---
title: "Cumulative"
author: Naveed Ahmad
format: html
execute: 
  echo: false
---

```{r}
#| message: false
#| warning: false
library(MASS)
library(tidyverse)
library(primer.data)
library(broom)
library(dplyr)
library(gt)
library(easystats)
library(marginaleffects)
library(ggplot2)
library(forcats)
library(janitor)
```

## Summary
<div style="text-align: justify">Political ideology may have significance for presidential approval in elections. In this study, using data from cooperative election survey 2020, I have attempted to answer the question; what is the average difference between political ideology groups and education for presidential approval of Trump in elections? However, my assumption of representatives of data may not hold good as data is collected from a group of people who may have different political ideology or activity then the whole population.I used cumulative regression model to predict presidential approval as function of political ideology and education. Predictions show that ideologically very conservative have the highest level of disapproval at around 5.45 with range from	5.34	to 5.55. However, with higher education and very liberal have the approval for presidential election of the President Trump.</div>

```{r}
x <- ces |>
  filter(year == 2020) |>
  select(approval, ideology, education) |>
  drop_na() |> 
  filter(! ideology %in% "Not Sure") |> 
  mutate(ideology = fct_drop(ideology))

tidy_approval <- x
print(tidy_approval)
if (!dir.exists("data")) dir.create("data")
write_rds(tidy_approval, "data/tidy_approval.rds")
tidy_approval <- read_rds("data/tidy_approval.rds")
```

## Plot showing approval and ideology

```{r}
x |>
  ggplot(aes(x = ideology, fill = approval)) +
  geom_bar() +
  labs(title = "Predidential approval and political ideology",
       subtitle = "Most people strongly disapprove",
       x = "political ideology",
       y = "count",
       fill = "approval of the president",
       caption = "Data from CES")
```

## Cumulative Logistic Regression Model

$$P(Y \leq k) = \frac{1}{1 + e^{-(\alpha_k - \beta_1 \cdot \text{ideology} - \beta_2 \cdot \text{education})}}$$

with $Y \sim \text{Ordinal}(\boldsymbol{\rho})$ where $\boldsymbol{\rho} = (\rho_1, \rho_2, \ldots, \rho_K)$ are derived from the cumulative probabilities above.


```{r}
fit_approval <- polr(approval ~ ideology + education, data = x)
```

## Coefficients

```{r}
#| cache: true
tidy_approval1 <- fit_approval |>
    tidy(conf.int = TRUE)
print(tidy_approval1)
```

## Fitted Cumulative Logistic Regression Model

We model the probability of presidential approval using an ordinal logistic regression (proportional odds model). The model is:

$$
\log\left( \frac{\Pr(\hat{Y} \leq k)}{\Pr(\hat{Y} > k)} \right)
= \theta_k 
- \left(
  0.615 \cdot \text{ideology}_{\text{Liberal}} +
  2.403 \cdot \text{ideology}_{\text{Moderate}} +
  4.464 \cdot \text{ideology}_{\text{Conservative}} +
  5.450 \cdot \text{ideology}_{\text{Very Conservative}} \right.
$$

$$
\left.
  -\; 0.096 \cdot \text{education}_{\text{High School}} 
  - 0.311 \cdot \text{education}_{\text{Some College}} 
  - 0.325 \cdot \text{education}_{\text{2-Year}} 
  - 0.521 \cdot \text{education}_{\text{4-Year}} 
  - 0.727 \cdot \text{education}_{\text{Post-Grad}} 
\right)
$$

where \( \hat{Y} \) is the predicted approval category, and the thresholds \( \theta_k \) correspond to the cumulative category boundaries:

$$
\begin{aligned}
\theta_1 &= 2.26 \quad \text{(Strongly Disapprove)} \\
\theta_2 &= 2.73 \quad \text{(Disapprove / Somewhat Disapprove)} \\
\theta_3 &= 2.82 \quad \text{(Neither)} \\
\theta_4 &= 3.93 \quad \text{(Approve / Somewhat Approve)}
\end{aligned}
$$

This is our Data Generating Mechanism

## Table showing coefficients and Conf.Interval
```{r}
 tidy_approval1|>
  select(term, estimate, conf.low, conf.high) |>
  gt() |>
  tab_header(
    title = "Coefficients and Confidence Intervals"
  ) |>
  fmt_number(
    columns = c(estimate, conf.low, conf.high),
    decimals = 2
  ) |>
  cols_label(
    estimate = "Estimate",
    conf.low = "95% CI Lower",
    conf.high = "95% CI Upper"
  ) |>
  tab_options(
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    column_labels.font.weight = "bold")
```


```{r}
# Split predictors and thresholds
predictors <- tidy_approval1 |> filter(!is.na(conf.low))
thresholds <- tidy_approval1 |> filter(is.na(conf.low))

# Plot with corrected column names
ggplot(predictors, aes(x = estimate, y = fct_reorder(term, estimate))) +
  geom_point(color = "#0072B2", size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "#0072B2") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Relationship of Ideology and Education on Presidential Approval",
    subtitle = "Estimates with 95% Confidence Intervals from Ordinal Logistic Regression",
    caption = "Data Source: CES 2020",
    x = "Estimate",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.caption = element_text(hjust = 0, face = "italic", size = 10))
```
