---
title: "Cumulative"
author: Naveed Ahmad
format: html
execute: 
  echo: false
---

```{r}
#| message: false
#| warning: false
library(MASS)
library(tidyverse)
library(primer.data)
library(broom)
library(dplyr)
library(gt)
library(easystats)
```

## Summary
<div style="text-align: justify">Political ideology may have significance for presidential approval in elections. In this study using data from cooperative election survey 2020, I have attempted to answer the question; what is the average difference between political ideology groups for presidential approval of Trump in elections? However, our assumption of representatives of data may not hold good as data is collected from a group of people who may have different political ideology or activity then the whole population in this study.</div>

```{r}
x <- ces |>
  filter(year == 2020) |>
  select(approval, ideology, education) |>
  drop_na() |> 
  filter(! ideology %in% "Not Sure") |> 
  mutate(ideology = fct_drop(ideology))

tidy_approval <- x
print(tidy_approval)
if (!dir.exists("data")) dir.create("data")
write_rds(tidy_approval, "data/tidy_approval.rds")
tidy_approval <- read_rds("data/tidy_approval.rds")
```

## Plot showing approval and ideology

```{r}
x |>
  ggplot(aes(x = ideology, fill = approval)) +
  geom_bar() +
  labs(title = "Predidential approval and political ideology",
       subtitle = "Most people strongly disapprove",
       x = "political ideology",
       y = "count",
       fill = "approval of the president",
       caption = "Data from CES")
```

## Cumulative Logistic Regression Model

$$P(Y \leq k) = \frac{1}{1 + e^{-(\alpha_k - \beta_1 \cdot \text{ideology} - \beta_2 \cdot \text{education})}}$$

with $Y \sim \text{Ordinal}(\boldsymbol{\rho})$ where $\boldsymbol{\rho} = (\rho_1, \rho_2, \ldots, \rho_K)$ are derived from the cumulative probabilities above.


```{r}
fit_approval <- polr(approval ~ ideology + education, data = x)
```

## Coefficients

```{r}
#| cache: true
clean_approval <- fit_approval |>
    tidy(conf.int = TRUE)
print(clean_approval)
```

## Fitted Cumulative Logistic Regression Model

We model the probability of presidential approval using an ordinal logistic regression (proportional odds model). The model is:

$$
\log\left( \frac{\Pr(\hat{Y} \leq k)}{\Pr(\hat{Y} > k)} \right)
= \theta_k 
- \left(
  0.615 \cdot \text{ideology}_{\text{Liberal}} +
  2.403 \cdot \text{ideology}_{\text{Moderate}} +
  4.464 \cdot \text{ideology}_{\text{Conservative}} +
  5.450 \cdot \text{ideology}_{\text{Very Conservative}} \right.
$$

$$
\left.
  -\; 0.096 \cdot \text{education}_{\text{High School}} 
  - 0.311 \cdot \text{education}_{\text{Some College}} 
  - 0.325 \cdot \text{education}_{\text{2-Year}} 
  - 0.521 \cdot \text{education}_{\text{4-Year}} 
  - 0.727 \cdot \text{education}_{\text{Post-Grad}} 
\right)
$$

where \( \hat{Y} \) is the predicted approval category, and the thresholds \( \theta_k \) correspond to the cumulative category boundaries:

$$
\begin{aligned}
\theta_1 &= 2.26 \quad \text{(Strongly Disapprove)} \\
\theta_2 &= 2.73 \quad \text{(Disapprove / Somewhat Disapprove)} \\
\theta_3 &= 2.82 \quad \text{(Neither)} \\
\theta_4 &= 3.93 \quad \text{(Approve / Somewhat Approve)}
\end{aligned}
$$

This is our Data Generating Mechanism

## Table showing coefficients and Conf.Interval
```{r}
 clean_approval|>
  select(term, estimate, conf.low, conf.high, coef.type) |>
  gt() |>
  tab_header(
    title = "Coefficients and Confidence Intervals"
  ) |>
  fmt_number(
    columns = c(estimate, conf.low, conf.high),
    decimals = 2
  ) |>
  cols_label(
    estimate = "Estimate",
    conf.low = "95% CI Lower",
    conf.high = "95% CI Upper",
    coef.type = "Coefficient Type"
  ) |>
  tab_options(
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    column_labels.font.weight = "bold")
```
